A netdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm

### 2025-08-01 paper.qmd
Addressing review comments (#1)
* Working on the review

* Working on the review

* Addressing final issues

* Adding final edits

* Improving section on exposures

* Adding final touches

---------

Co-authored-by: aoliveram <anibal.olivera.m@gmail.com>

---
title: "A netdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm"
format: pdf
keep-tex: true
bibliography: bibliography.bib
bibliographystyle: apa
link-citations: true
author: 
  - name: George G. Vega Yon
    affiliation:
      - name: University of Utah
    email: george.vegayon@utah.edu
  - name: Aníbal L. Olivera M.
    affiliation:
      name: Universidad del Desarrollo
  - name: Thomas W. Valente
    affiliation:
      name: University of Southern California
# include-in-header: header.tex
---

@@

This work was supported by the National Institute of Drug Abuse, the National Institutes of Health (NIH), grant number R01-DA051843.

@@

The netdiffuseR R package [@netdiffuser_1; @netdiffuser_2] was first released to CRAN on February 18, 2016. The package contains a collection of algorithms and methods first envisioned by Dr. Valente and released in his book [@valente_book_1999] as GAUSS code. Since its release in 2016, the package has been downloaded over 45 thousand times and has about 400 downloads a month, with an active community growing.

@@

@@

Another thing to note are the warnings. The R programming language (like most do) distinguishes between decimal ("`numeric`") and integer numbers. In this case, the function is throwing a warning about columns beign coerced from numeric into integer types. These are mostly informative, and if the user rather skipping these, they can use the `suppressWarnings()` function in R. We will see more warnings as we execute R code and will pause at some points to discuss further about what these mean to the user.

@@

The function prints a couple of messages: The first, a warning about the `value` parameter beign swithced to `TRUE`, indicates that, since we are using the structural equivalence alternative graph, instead of treating the network as unweighted (0/1-valued ties), the function will compute exposures using weights. The second is a more technical internal message that is describing a change in the `Matrix` R package: users should not worry about this particular message.[^dgcmatrix]

@@

```{r}
#| label: visualizing-exposures
# Visualizing exposure at time 5
cbind(
  "Cohesive"       = exp_cohesive[, 5],
  "SE"             = exp_se[, 5],
  "Attr. weighted" = exp_age[, 5]
) |> head(n = 10)
```

@@

The warnings in `exposure()` indicate that (a) the passed graph is going to be avaluated as a valued (weighted) network and (b) the same graph is going to be used to compute exposure in all time points (so it is static). The next code chunk gives a glimpse of the different exposure values we have just calculated:

```{r}
#| label: showing-all-exposures
# Visualizing exposure at time 5
cbind(
  "Cohesive"       = exp_cohesive[, 5],
  "SE"             = exp_se[, 5],
  "Attr. weighted" = exp_age[, 5],
  "Homophilic"     = exp_age_homo[, 5]
) |> head(10)
```

@@

>>>>

---
title: "A netdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm"
format: pdf
keep-tex: true
bibliography: bibliography.bib
bibliographystyle: apa
link-citations: true
author: 
  - name: George G. Vega Yon
    corresponding: true
    email: george.vegayon@utah.edu
    affiliation:
      - name: University of Utah
  - name: Aníbal L. Olivera M.
    affiliation:
      name: Universidad del Desarrollo
  - name: Thomas W. Valente
    affiliation:
      name: University of Southern California
embed-resources: true
---

@@

@@

The netdiffuseR R package [@valente_yon_2020; @netdiffuser_2] was first released to CRAN on February 18, 2016. The package contains a collection of algorithms and methods first envisioned by Dr. Valente and released in his book [@valenteNetworkModelsDiffusion1995] as GAUSS code. Since its release in 2016, the package has been downloaded over 45 thousand times and has about 400 downloads a month, with an active community growing.

@@

This tutorial has been developed using **R** version 4.5.1 [@R] and the program **quarto** version 1.6.42 [@Allaire_Quarto_2025]. Readers can find the quarto document as well as all required material for reproducing the paper in <https://github.com/gvegayon/netdiffuser-connections>.

@@

Another thing to note are the warnings. The R programming language (like most do) distinguishes between decimal ("`numeric`") and integer numbers. In this case, the function is throwing a warning about columns beign coerced from numeric into integer types. These are mostly informative, and the user can skipping these with the `suppressWarnings()` function in R. We will see more warnings as we execute R code and will pause at some points to discuss what these mean to the user.

@@

The function prints a couple of messages: The first, a warning about the `value` parameter being swithced to `TRUE`, indicates that, since we are using the structural equivalence alternative graph, instead of treating the network as unweighted (0/1-valued ties), the function will compute exposures using weights. The second is a more technical internal message that is describing a change in the `Matrix` R package: users should not worry about this particular message.[^dgcmatrix]

@@

```{r}
#| label: visualizing-exposures
# Visualizing exposure at time 5
cbind(
  "Cohesion"       = exp_cohesive[, 5],
  "SE"             = exp_se[, 5],
  "Attr. weighted" = exp_age[, 5]
) |> head(n = 10)
```

@@

The warnings in `exposure()` indicate that (a) the passed graph is going to be evaluated as a valued (weighted) network and (b) the same graph is going to be used to compute exposure in all time points (so it is static). The next code chunk gives a glimpse of the different exposure values we have just calculated:

```{r}
#| label: showing-all-exposures
# Visualizing exposure at time 5
cbind(
  "Cohesion"       = exp_cohesive[, 5],
  "SE"             = exp_se[, 5],
  "Attr. weighted" = exp_age[, 5],
  "Homophilic"     = exp_age_homo[, 5]
) |> head(10)
```

@@

This work was supported by the National Institute of Drug Abuse, the National Institutes of Health (NIH), grant number R01-DA051843. We would also like to thank an anonymous reviewer for their comments and suggestions that improved this paper.
